partial: An R Package for Creating Partial Dependence Plots

### 2016-08-09 manuscript/greenwell.tex
minor edits

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a traditional random forest to the entire data set using the following code snippet:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

>>>>

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.
